Gravitational waves from colliding black holes

### 2017-05-02 Report/Draft/draft.tex
Report latest

\section{Detector response simulations}
In this section we describe the process of inserting signals into simulated detector noise, and then perform a full parameter estimation.
\subsection{Signal extraction}
Data whitening - basically LOSC stuff, how we go from raw data to a signal - move this to intro?
\subsection{Inference pipeline}
A large number of software injections were going to be necessary to generate any interesting results due to the large size and complexity of the parameter space as revealed by the match results. So it was key that they could be performed efficiently and in an organised way. This required the construction of an inference pipeline which formed the bulk of the computational work of the project. This was ultimately one bash script that would create a new folder for each run, generate the injection file and run the inference MCMC.  It also saves the injected parameters as a python dictionary, plots all posteriors overlaid with injected parameters, plots the injected waveform, and the recovered (maximum posterior, or \textit{MAP} waveform) and injected waveforms overlaid on the whitened detector strain data, and return matches between the MAP and injected waveforms. This meant that all the relevant analysis and data processing needed for each inference run was fully automated. A flow chart of this process is shown in Fig. 12. For the MCMC parameters, in order to acheive a balance between computation time and result accuracy, 5,000 samplers were used and 12,000 iterations, and the burn-in process was skipped. These parameters were kept the same for all software injections.
This project work was also split between a number of different LIGO clusters, as well as work on several different local PCs, and so the pipeline along with all the relevant match and precession scripts were maintained as part of a GitHub repository which meant that the codes could be developed effectively and without conflict.
	\includegraphics[scale=0.55]{fig13.png}
	\caption{Flow chart of the inference pipeline. The ellipses represent data files that were generated during the pipeline and stored in case further analysis was necessary. The boxes at the bottom represent the final output of the pipeline.}
\subsection{Inference results}
In this section we present the results from a range of inference runs, attempt to assess the effectiveness of spin estimation and compare the inference results with the match results to evaluate how effective the match process is as a predictor of inference accuracy. For all injections we set $S_{1\perp}=0.9$ and $|\vec{S_2}|=0$ with $\chi_p=0.9$, and cover a uniform range of inclinations. We set the phase to be $\phi=0.$, and the polarisation angle $\psi=0.8$, so the signals are of mixed polarisation, but slightly cross dominated. The match results indicate that cross polarised signals generally have a higher spin resolution, but we want to avoid the signal drop at $\iota=\pi/2$, which is why a mixed but cross dominated polarisation angle was selected. The source location and coalesence time were also kept fixed for all injections. We also only examine high mass ratio binaries, as the match results indicate that precession is not significant in equal mass ratio binaries. Out of a total of ****** software injections that were performed, only **** instances where the injected $\chi_p$ value fell within the 90\% confidence interval. In Figs 14 and 15 we compare posteriors for two inferences, where $\chi_p$ was excluded and where it was recovered moderately well, and we refer to these as A and B respectively. Even in the cases where the injected value was within the 90\% confidence intervals, it was always closer to the upper bound than to the mean, and a mean $\chi_p$ greater than 0.65 was never found.
	\includegraphics[scale=1]{fig14.png}
	\caption{Posteriors for an injection with $m1=50$, $m2=15$, $\iota=2.51$, $\phi=0$, $\psi=0.8$ at a distance of 200Mpc. Prior distributions are included for the derived spin parameters $\chi_p$ and $\chi_{eff}$, which result from uniform spin magnitude and isotropic orientation distributions.}
	\includegraphics[scale=1]{fig15.png}
	\caption{Posteriors for an injection with the same parameters as in Fig. 14, expect here with an inclination of $\iota=0.63$}

@@

\begin{figure}[h]
	\includegraphics[scale=0.3]{fig16.png}
	\caption{Strain plots for both H1 and L1 detectors. The top two graphs show the strain data corresponding to the results in Fig. 14, the bottom two graphs correspond to Fig. 15. The green and red lines show the whitened strain data, the black line is the MAP waveform and the injected waveform is shown in blue.}

@@

If two signals are degenerate for the majority of the inspiral and merger, and only during the end of the merger and ringdown do they differ, it will be extremely difficult for LIGO detectors to differentiate between them, as the very short time period where there is any mismatch will be swamped by noise. Returning to Fig. 14, we can see that even though the match is strong, there are significant inaccuracies on $\chi_p$ and $\iota$. So in results A and B we have examples of each inaccuracy as described earlier - one down to poor signal recovery and one down to an intrinsic degeneracy.
\subsection{Inclination and precession}
Next we examine the effect of inclination on $\chi_p$ recovery. Software injections were performed for a range of inclinations at two distances, 200Mpc and 800Mpc, which simulate a strong and weak signal respectively.
\subsection{Impact of Virgo}
A look at how Virgo will influence PE, especially chi p estimation.
\section{Conclusions}
Summary of results on PE and estimation of chi p, and a prospect on Virgo's impact.
\section{Reflective Statement}
I now need a therapist

>>>>

\section{Detector response simulations}
In this section we describe the process of inserting signals into simulated detector noise, and then perform a full parameter estimation.
\subsection{Signal extraction}
Data whitening - basically LOSC stuff, how we go from raw data to a signal - move this to intro?
\subsection{Inference pipeline}
A large number of software injections were going to be necessary to generate any interesting results due to the large size and complexity of the parameter space as revealed by the match results. So it was key that they could be performed efficiently and in an organised way. This required the construction of an inference pipeline which formed the bulk of the computational work of the project. This was ultimately one bash script that would create a new folder for each run, generate the injection file and run the inference MCMC.  It also saves the injected parameters as a python dictionary, plots all posteriors overlaid with injected parameters, plots the injected waveform, and the recovered (maximum posterior, or \textit{MAP} waveform) and injected waveforms overlaid on the whitened detector strain data, and return matches between the MAP and injected waveforms. This meant that all the relevant analysis and data processing needed for each inference run was fully automated. A flow chart of this process is shown in Fig. 12. For the MCMC parameters, in order to acheive a balance between computation time and result accuracy, 5,000 samplers were used and 12,000 iterations, and the burn-in process was skipped. These parameters were kept the same for all software injections.
This project work was also split between a number of different LIGO clusters, as well as work on several different local PCs, and so the pipeline along with all the relevant match and precession scripts were maintained as part of a GitHub repository which meant that the codes could be developed effectively and without conflict.
\begin{figure}
	\includegraphics[scale=0.55]{fig13.png}
	\centering
	\caption{Flow chart of the inference pipeline. The ellipses represent data files that were generated during the pipeline and stored in case further analysis was necessary. The boxes at the bottom represent the final output of the pipeline.}
	\centering
\end{figure}
\subsection{Inference results}
In this section we present the results from a range of inference runs, attempt to assess the effectiveness of spin estimation and compare the inference results with the match results to evaluate how effective the match process is as a predictor of inference accuracy. For all injections we set $S_{1\perp}=0.9$ and $|\vec{S_2}|=0$ with $\chi_p=0.9$, and cover a uniform range of inclinations. We set the phase to be $\phi=0.$, and the polarisation angle $\psi=0.8$, so the signals are of mixed polarisation, but slightly cross dominated. The match results indicate that cross polarised signals generally have a higher spin resolution, but we want to avoid the signal drop at $\iota=\pi/2$, which is why a mixed but cross dominated polarisation angle was selected. The source location and coalesence time were also kept fixed for all injections. We also only examine high mass ratio binaries, as the match results indicate that precession is not significant in equal mass ratio binaries. Out of a total of ****** software injections that were performed, only **** instances where the injected $\chi_p$ value fell within the 90\% confidence interval. In Figs 14 and 15 we compare posteriors for two inferences, where $\chi_p$ was excluded and where it was recovered moderately well, and we refer to these as A and B respectively. Even in the cases where the injected value was within the 90\% confidence intervals, it was always closer to the upper bound than to the mean, and a mean $\chi_p$ greater than 0.65 was never found.

\begin{figure}
	\includegraphics[scale=1]{fig14.png}
	\centering
	\caption{Posteriors for an injection with $m1=50$, $m2=15$, $\iota=2.51$, $\phi=0$, $\psi=0.8$ at a distance of 200Mpc. Prior distributions are included for the derived spin parameters $\chi_p$ and $\chi_{eff}$, which result from uniform spin magnitude and isotropic orientation distributions.}
	\centering
\end{figure}
\begin{figure}
	\includegraphics[scale=1]{fig15.png}
	\centering
	\caption{Posteriors for an injection with the same parameters as in Fig. 14, expect here with an inclination of $\iota=0.63$}
	\centering
\end{figure}

@@

\begin{figure}[h]
	\includegraphics[scale=0.3]{fig16.png}
	\centering
	\caption{Strain plots for both H1 and L1 detectors. The top two graphs show the strain data corresponding to the results in Fig. 14, the bottom two graphs correspond to Fig. 15. The green and red lines show the whitened strain data, the black line is the MAP waveform and the injected waveform is shown in blue.}
	\centering
\end{figure}

@@

If two signals are degenerate for the majority of the inspiral and merger, and only during the end of the merger and ringdown do they differ, it will be extremely difficult for LIGO detectors to differentiate between them, as the very short time period where there is any mismatch will be swamped by noise. Returning to Fig. 14, we can see that even though the match is strong, there are significant inaccuracies on $\chi_p$ and $\iota$. So in A and B we have examples of each inaccuracy as described earlier - one down to poor signal recovery and one down to an intrinsic degeneracy. In Figs 17 and 18, results are shown for case C, which was selected as its waveform had the strongest visible precessive effects in both H1 and L1 detector frames.

From Fig. 18 it can be seen that the amplitude of the signal noticeably drops towards the end of the inspiral and just before the merger, implying that this would be an ideal signal to recover a high value of $\chi_p$ from. The posteriors shown in Fig. 17 are mostly accurate, with the majority of injected parameters lying inside the 90\% confidence intervals, however once again $\chi_p$ is excluded, and significantly underestimated. The match values shown in Fig. 18 are strong, with both detectors giving a match of 0.96 between the injected and MAP waveforms. This is further evidence of an intrinsic degeneracy that prevents high $\chi_p$ MAP values from being found. For all posteriors shown, both $\phi$ and $\psi$ are broadly uniform, and in the case where they are not uniform, they are largely inaccurate. However in section 2 it was observed that $\psi$ and especially $\phi$ have a significant effect on the structure of the parameter space, and the match between precessing and non-precessing waveforms. One possible explanation for this uniformity is that for each stochastically selected phase and polarisation, the MCMC algorithm readily finds a good match between the injected signal and the strain data simply by tweaking the other remaining parameters slightly. If this is the case, since there are a larger number of waveforms in the template bank with lower values of $\chi_p$ (as shown in the prior distributions), if these waveforms can still match the signal well just by slightly changing inclination, source location and masses, we would expect the MCMC algorithm to favour lower values of $\chi_p$ but with slightly larger uncertanties on the other posteriors. This explanation is exemplified by the waveforms in Fig. 8, where it is shown that the magnitude of the precessive effect is not necessarily proportional to $\chi_p$. It is also interesting to note that $\chi_{eff}$ parameters are generally relatively accurate, which implies it is mainly spin magnitude that is being underestimated, but the orientations are recovered accurately.
\begin{figure}
	\includegraphics[scale=1]{fig17.png}
	\centering
	\caption{Posteriors for injection C, which has the same parameters as in Fig. 14, but with an inclination of $\iota=1.26$ and at a distance of 800Mpc.}
	\centering
\end{figure}
\begin{figure}
	\includegraphics[scale=0.9]{fig18.png}
	\centering
	\caption{Strain plot of injection C, a waveform with strong precessive modulation.}
	\centering
\end{figure}
\subsection{Inclination and precession}
Next we examine the effect of inclination on $\chi_p$ recovery. Software injections were performed for a range of inclinations at two distances, 200Mpc and 800Mpc, which simulate a strong and weak signal respectively.
\subsection{Impact of Virgo}
A look at how Virgo will influence PE, especially chi p estimation.
\section{Conclusions}
Summary of results on PE and estimation of chi p, and a prospect on Virgo's impact.
\section{Reflective Statement}
I now need a therapist
